Stage-Splitting a Modal Language

### 2014-10-01 paper/example.tex
changes from Kayvon and David's comments; also some color

fun qSelect (l : list, k : int) : int = 
  case l of
    Empty => 0
  | Cons (h,t) => 
      let (left,right,n) = partition h t in
        case compare k n of
          LT => qSelect (left, k)
        | EQ => h
        | GT => qSelect (right, k-n-1)
\end{lstlisting}
\caption{Unstaged implementation of quickselect.}
\label{fig:quickselect}
\end{subfigure}%
\begin{subfigure}{0.5\textwidth}
\begin{lstlisting} 
datatype list = Empty | Cons of int * list
fun partition (p : int, l : list) = ...

fun qsStaged (l : list, k : $int) : $int = 
  case l of
    1`Empty` => next {2`0`}
  | Cons (h,t) => 
      let (left,right,n) = partition h t in
      next{
        let n = hold{n} in
          case compare prev{k} n of
            LT => prev {qsStaged (left k)}
          | EQ => hold {h}
          | GT => prev {qsStaged (right, 
                             next{prev{k}-n-1)}}
      }	
\end{lstlisting}
\caption{Staged implementation of quickselect in \lang.}
\label{fig:qsstaged}
\end{subfigure}
\begin{subfigure}{0.5\textwidth}
\begin{lstlisting}
datatype list = Empty | Cons of int * list
fun partition (p : int, l : list) = ...

@@

fun qSelect1 (l : list) : tree =
  case l of
    Empty => Leaf
  | Cons (h,t) => 
      let (left,right,n) = partition h t in
      Branch (n, h, qSelect1 left, qSelect1 right)

fun qSelect2 (p : tree, k : int) : int = 
  case unroll p of
    Leaf => 0
  | Branch (n,h,p1,p2) => 
      case compare k n of
        LT => qSelect2 (p1,k)
      | EQ => h
      | GT => qSelect2 (p2,k-n-1)
\end{lstlisting}
\caption{Split (two-phase) implementation of quickselect.}
\label{fig:qssplit}
\end{subfigure}
\caption{Caption place holder}
\end{figure*}

@@

One notable property of quickselect is that the partitioning of the list, even through its recursive calls, depends only on the values of list elements, not on the value of $k$ 
\footnote{There is a control-dependence on $k$ -- that is $k$ determines what splits are made but not how they are made -- we ignore this for now.}.  Therefore, if quickselect was to be evaluated many times on the same list, this independence admits a familiar optimization where the algorithm is split into two functions: one which precomputes \emph{all partitions} of the input list needed for any value of $k$ (resulting in a binary search tree---see \texttt{qSelect1} in \ref{fig:qssplit}),
and another which, given $k$, uses the tree to perform accelerated lookup of the $k$-th element (essentially performing binary search of a sorted list---see \texttt{qSelect2}).

The benefits are substantial: for a randomly ordered list of length $n$, 
the original \texttt{qSelect} takes expected $O(n)$ time per lookup, 
whereas \texttt{qSelect2} takes $O(\log n)$ time.
Since \texttt{qSelect1} does not depend on $k$, its $O(n\log n)$ cost can be amortized over many lookups.
%One advantage of splitting the algorithm into two phases is that the search tree can be constructed for a list once
%and then used to accelerate many future lookups for any value of $k$.

@@

Each construct in the body of quickselect is now associated with a stage via an interleaving of $\next$ and $\prev$ blocks.  
Specifically, $\next$ occurs in a stage \bbone\ context and indicates that the contents of its block are stage \bbtwo, 
whereas $\prev$ occurs in a stage \bbtwo\ context and indicates that the contents of its block is stage \bbone.
(We adapt the convention that the top-level context is stage \bbone.)
The output type of a $\next$ block is the $\fut$'d version of the type of its stage \bbtwo\ contents.  
For example, \verb|next{0}| from above has the type $\fut\rmint$.
Correspondingly, $\prev$ requires that its stage \bbone\ contents have a $\fut$ type, and it eliminates the wrapper.
For example, \verb|prev{k}| from above has type $\rmint$ at stage \bbtwo, since $k$ is bound to a $\fut\rmint$ at stage \bbone.
These type restrictions essentially enforce that ``later stage content" is always treated hygienically at stage \bbone,
which is necessary to admit a properly staged implementation.

>>>>

fun qSelect (l : list, k : int) : int = 
  case l of
    Empty => 0
  | Cons (h,t) => 
      let (left,right,n) = partition h t in
        case compare k n of
          LT => qSelect (left, k)
        | EQ => h
        | GT => qSelect (right, k-n-1)
\end{lstlisting}
\caption{Unstaged implementation of quickselect.}
\label{fig:quickselect}
\end{subfigure}%
\begin{subfigure}{0.5\textwidth}
\begin{lstlisting} 
1`datatype list = Empty | Cons of int * list
fun partition (p : int, l : list) = ...

fun qsStaged (l : list, k : $`2`int`1`) : $`2`int`1` = 
  case l of
    Empty => next {`2`0`1`}
  | Cons (h,t) => 
      let (left,right,n) = partition h t in
      next{
        `2`let n = hold{`1`n`2`} in
          case compare prev{`1`k`2`} n of
            LT => prev {`1`qsStaged (left k)`2`}
          | EQ => hold {`1`h`2`}
          | GT => prev {`1`qsStaged (right, 
                             next{prev{k}-n-1)}`2`}`1`
      }`
\end{lstlisting}
\caption{Staged implementation of quickselect in \lang.}
\label{fig:qsstaged}
\end{subfigure}
\begin{subfigure}{0.5\textwidth}
\begin{lstlisting}
1`datatype list = Empty | Cons of int * list
fun partition (p : int, l : list) = ...`

@@

1`fun qSelect1 (l : list) : tree =
  case l of
    Empty => Leaf
  | Cons (h,t) => 
      let (left,right,n) = partition h t in
      Branch (n, h, qSelect1 left, qSelect1 right)`

2`fun qSelect2 (p : tree, k : int) : int = 
  case unroll p of
    Leaf => 0
  | Branch (n,h,p1,p2) => 
      case compare k n of
        LT => qSelect2 (p1,k)
      | EQ => h
      | GT => qSelect2 (p2,k-n-1)`
\end{lstlisting}
\caption{Split (two-phase) implementation of quickselect.}
\label{fig:qssplit}
\end{subfigure}
\caption{Caption place holder}
\end{figure*}

@@

One notable property of quickselect is that the partitioning of the list, even through its recursive calls, depends only on the values of list elements, not on the value of $k$ 
\footnote{There is a control-dependence on $k$ -- that is $k$ determines what splits are made but not how they are made -- we ignore this for now.}.  Therefore, if quickselect were to be evaluated many times on the same list, this independence admits a familiar optimization where the algorithm is split into two functions: one which precomputes \emph{all partitions} of the input list needed for any value of $k$ (resulting in a binary search tree---see \texttt{qSelect1} in \ref{fig:qssplit}),
and another which, given $k$, uses the tree to perform accelerated lookup of the $k$-th element (essentially performing binary search of a sorted list---see \texttt{qSelect2}).  
Equationally, we can represent this relationship as 
\[
\mathtt{qSelect(l,k)} \equiv \mathtt{qSelect2(qSelect1(l),k)}
\]

The benefits of this decomposition are substantial: for a randomly ordered list of length $n$, 
the original \texttt{qSelect} takes expected $O(n)$ time per lookup, 
whereas \texttt{qSelect2} takes $O(\log n)$ time.
Since \texttt{qSelect1} does not depend on $k$, its $O(n\log n)$ cost can be amortized over many lookups.
%One advantage of splitting the algorithm into two phases is that the search tree can be constructed for a list once
%and then used to accelerate many future lookups for any value of $k$.

@@

Each construct in the body of quickselect is now associated with a stage via an interleaving of $\next$ and $\prev$ blocks.  
Specifically, $\next$ occurs in a stage \bbone\ context and indicates that the contents of its block are stage \bbtwo, 
whereas $\prev$ occurs in a stage \bbtwo\ context and indicates that the contents of its block is stage \bbone.
(We adopt the convention that the top-level context is stage \bbone.)
The output type of a $\next$ block is the $\fut$'d version of the type of its stage \bbtwo\ contents.  
For example, \verb|next{0}| from above has the type $\fut\rmint$.
Correspondingly, $\prev$ requires that its stage \bbone\ contents have a $\fut$ type, and it eliminates the wrapper.
For example, \verb|prev{k}| from above has type $\rmint$ at stage \bbtwo, since $k$ is bound to a $\fut\rmint$ at stage \bbone.
These type restrictions essentially enforce that ``later stage content" is always treated hygienically at stage \bbone,
which is necessary to admit a properly staged implementation.

### 2014-10-01 paper/paper.tex
changes from Kayvon and David's comments; also some color

\lstset{language=ML,
        columns=fullflexible,
        keepspaces=true,
        basicstyle=\ttfamily,
        tabsize=4,
        escapeinside={"*}{*"},
        moredelim=*[is][\color{red}]{1`}{`},
        moredelim=*[is][\color{blue}]{2`}{`},
        literate={|->}{{\ $\mapsto$\ }}1
                 {xhat}{{$\hat{\texttt{x}}$}}1
                 {yhat}{{$\hat{\texttt{y}}$}}1
                 {zhat}{{$\hat{\texttt{z}}$}}1
                 {murec}{{$\mu$}}1
                 {alpha}{{$\alpha$}}1
                 {beta}{{$\beta$}}1
                 {$}{{$\fut$}}1}

>>>>

\lstset{%language=ML,
        columns=fullflexible,
        keepspaces=true,
        basicstyle=\ttfamily,
        tabsize=4,
        escapeinside={"*}{*"},
        moredelim=*[is][\color{red}]{1`}{`},
        moredelim=*[is][\color{blue}]{2`}{`},
        literate={|->}{{\ $\mapsto$\ }}1
                 {xhat}{{$\hat{\texttt{x}}$}}1
                 {yhat}{{$\hat{\texttt{y}}$}}1
                 {zhat}{{$\hat{\texttt{z}}$}}1
                 {murec}{{$\mu$}}1
                 {alpha}{{$\alpha$}}1
                 {beta}{{$\beta$}}1
                 {$}{{$\fut$}}1}

### 2014-10-01 paper/semantics.tex
fix spacing in listings
changes from Kayvon and David's comments; also some color

\subsubsection{Non-Duplicating Stage~\bbone\ Evaluation}

To gain intuition about the challenges of implementing this staged dynamic
semantics, consider the program:
\begin{lstlisting}
#2 (next {f 20}, 2+3)
\end{lstlisting}
This is a stage~\bbone\ expression of type $\rmint$; the pair inside it is a
stage~\bbone\ expression of type $(\fut\rmint)\times\rmint$; and ${\tt f} :
\rmint \to \rmint$ is an expensive stage~\bbtwo\ function which we would like
to avoid evaluating. A conventional call-by-value semantics would fully
evaluate both components of the pair before projecting the second component.
The problem is that, while \verb|next {f 20}| is not a value (in the sense that
additional stage-\bbtwo\ computation steps are necessary to produce a numeral),
evaluating the contents of \verb|next| cannot occur as part of stage~\bbone\
evaluation.

Intuitively, the solution is to designate \verb|next {f 20}| as a value
\emph{in stage \bbone}, even though it requires additional evaluation in stage
\bbtwo. Therefore, the pair evaluates to
\begin{lstlisting}
(next {f 20}, 5)
\end{lstlisting}
and the projection, in turn, immediately evaluates to \verb|5|.

Now consider a more complex example where stage~\bbone\ evaluation must
substitute such a incompletely-evaluated expression. The following
stage-\bbtwo\ term has type $\rmint$:
\begin{lstlisting} 
prev{
  let x = (next {f 20}, 3+4) in
  next{ prev{#1 x} * prev{#1 x} * hold{#2 x} }
}
\end{lstlisting}
As with the simpler example, this term does not fully reduce at stage~\bbone, 
because it depends on the value of \verb|f 20|, which is not reduced until stage \bbtwo.

Again treating \verb|(next {f 20}, 7)| as a value during stage~\bbone, we
substitute it for the three occurrences of \verb|x| in the body of the
\verb|let| expression, yielding
\begin{lstlisting} 
prev{
  next{ prev{#1 (next {f 20}, 7)} 
      * prev{#1 (next {f 20}, 7)} 
      * hold{#2 (next {f 20}, 7)} }
}
\end{lstlisting}
To then evaluate the large $\next$, we must first search it for stage~\bbone\ subterms to reduce in place.
This search process is called {\em speculation}.
In our example this means reducing all three projections to give,
\begin{lstlisting} 
prev{
  next{ 
    prev{next {f 20}} * prev{next {f 20}} * 7 
  }
}
\end{lstlisting}
Since all stage~\bbone\ subterms are reduced, the $\prev$s and $\next$s cancel to yield the final residual:
\begin{lstlisting} 
(f 20) * (f 20) * 7
\end{lstlisting}
But now stage \bbtwo\ must compute the expensive function call \verb|f 20| twice! 
There are some systems, typically called {\em metaprogramming}, 
where this duplication would be considered intended behavior.
For our applications, we find the duplication undesirable and avoid it.

@@

More precisely, when evaluating the \verb|next|, we construct an explicit
substitution $\mathtt{[\hat y\mapsto f~20]}$ binding its old contents to the
fresh variable $\mathtt{\hat y}$. This substitution is placed at the top of the
containing $\prev$ block:
\begin{lstlisting} 
prev {
[yhat|->f 20]
  let x = (next{yhat}, 7) in
  next{prev{#1 x} * prev{#1 x} *  hold{#2 x}}
}
\end{lstlisting}
%As a convention, we render the new variable with a %stylish and fashionable
%hat.  
We proceed with stage~\bbone\ evaluation, which now duplicates the variable
$\mathtt{\hat y}$ rather than the expression \verb|f 20|.
\begin{lstlisting} 
prev {
[yhat|->f 20]
  next{
    prev{#1 (next {yhat}, 7)} * 
    prev{#1 (next {yhat}, 7)} *
    hold{#2 (next {yhat}, 7)}
  }
}
\end{lstlisting}
We then speculate to reduce the projections and cancel the $\next$s and $\prev$s, giving:
\begin{lstlisting} 
prev {
[yhat|->f 20]
    next{ yhat * yhat * 7 }
}
\end{lstlisting}
Once again, we lift the results of speculation into a substitution:
\begin{lstlisting} 
prev {
[yhat|->f 20]
[zhat|->yhat*yhat*7]
    next{ zhat }
}
\end{lstlisting}
Finally, when canceling the last $\next$ and $\prev$, we must {\em reify} the contained substitutions into let statements, yielding
\begin{lstlisting} 
let yhat = f 20 in
let zhat = yhat * yhat * 7 in zhat
\end{lstlisting}

@@

The dynamics for \lang\ implements the stage \bbone\ evaluation algorithm
described above, using three judgments: $\redonesym$ (evaluation), $\redtwosym$
(speculation), and $\reifysym$ (reification of substitutions into let
bindings).

\lang\ creates two different kinds of expressions which
cannot be evaluated further at stage \bbone:
\begin{itemize}
\item Partial values ($\pvalsym$s) are stage \bbone\ terms that have been fully
evaluated, but which may contain stage-\bbtwo\ variables wrapped in $\next$
blocks.  In the example above, 
\verb|(|$\next \{\mathtt{\hat y}\}$\verb|,7)|.

\item Residuals ($\ressym$es) are \langTwo\ terms---stage \bbtwo\ terms whose
stage \bbone\ subexpressions have all been fully evaluated. In the example
above,
\verb|(1+2)| and \verb|(let z = y*y*7 in z)|.

%\item Values ($\valsym$s) are \langTwo\ terms which are fully evaluated; these
%are the results of a computation after both stages have been completed. In the
%example above, the term evaluates to \verb|63|.
\end{itemize}

@@

At $\next~e$, in contrast, $\redonesym$ must search through $e$ for any stage
\bbone\ subexpressions, and evaluate them.  (See \ref{fig:diaSemanticsNP}.)
This search process, called \emph{speculation}, is implemented by the
$\redtwosym$ judgment, which takes a stage \bbtwo\ term to a residual.

For the most part (\ref{fig:diaSemanticsSpec}), speculation simply recurs into
all subexpressions; at $\prev$, however, it resumes $\redonesym$ evaluation,
which produces a future context $\xi$ and (by canonical forms) an expression
$\next\{\hat y\}$. We then reify $\xi$ using $\reifysym$ into a series of let
bindings enclosing $\mathtt{\hat y}$.

@@

Within speculation lies a subtle---if perhaps unintuitive---feature.  
Observe that speculation will traverse into both branches of a stage-\bbtwo\ {\tt if} or {\tt case} 
statement in its search for stage~\bbone\ code. 
Thus the evaluation of that stage~\bbone\ code will occur {\em regardless of the eventual value of the predicate},
and so a term like 
\begin{lstlisting} 
next{
  if true 
  then hold{1+2} 
  else prev{spin() (* loops forever *)}
}
\end{lstlisting}
will fail to evaluate at stage \bbone.
This behavior is why the judgment is named ``speculation."
Although it may seem undesirable, this speculative behavior is a deliberate feature.
Indeed it is critical to the quickselect example, 
wherein all three branches of the $n$-$k$ comparison are evaluated at stage~\bbone.
As always, it is the programmer's responsibility to ensure that her staging annotations 
do not cause a program to loop forever.

>>>>

\subsubsection{Stage~\bbone\ Evaluation}

To gain intuition about the challenges of implementing this staged dynamic
semantics, consider the program:
\begin{lstlisting}
1`#2 (next {`2`f 20`1`}, 2+3)`
\end{lstlisting}
This is a stage~\bbone\ expression of type $\rmint~@~\bbone$; the pair inside it is a
stage~\bbone\ expression of type $(\fut\rmint)\times\rmint~@~\bbone$; and ${\tt f} :
\rmint \to \rmint~@~\bbtwo$ is an expensive stage~\bbtwo\ function which we would like
to avoid evaluating. A conventional call-by-value semantics would fully
evaluate both components of the pair before projecting the second component.
The problem is that, while \verb|next {f 20}| is not a value (in the sense that
additional stage-\bbtwo\ computation steps are necessary to produce a numeral),
evaluating the contents of \verb|next| cannot occur as part of stage~\bbone\
evaluation.

Intuitively, the solution is to designate \verb|next {f 20}| as a value
\emph{in stage \bbone}, even though it requires additional evaluation in stage
\bbtwo. Therefore, the pair evaluates to
\begin{lstlisting}
1`(next {`2`f 20`1`}, 5)`
\end{lstlisting}
and the projection, in turn, immediately evaluates to \verb|5|.

Now consider a more complex example where stage~\bbone\ evaluation must
substitute such an incompletely-evaluated expression. The following
stage-\bbtwo\ term has type $\rmint$:
\begin{lstlisting} 
2`prev{`
  1`let x = (next {`2`f 20`1`}, 3+4) in
  next{` 2`prev{`1`#1 x`2`} * prev{`1`#1 x`2`} * hold{`1`#2 x`2`}` 1`}`
2`}`
\end{lstlisting}
As with the simpler example, this term does not fully reduce at stage~\bbone, 
because it depends on the value of \verb|f 20|, which is not reduced until stage \bbtwo.

Again treating \verb|(next {f 20}, 7)| as a value during stage~\bbone, we
substitute it for the three occurrences of \verb|x| in the body of the
\verb|let| expression, yielding
\begin{lstlisting} 
2`prev{`
  1`next{` 2`prev{`1`#1 (next {`2`f 20`1`}, 7)`2`} 
      * prev{`1`#1 (next {`2`f 20`1`}, 7)`2`} 
      * hold{`1`#2 (next {`2`f 20`1`}, 7)`2`} `1`}`2`
}`
\end{lstlisting}
The outermost $\next$ here is not yet a value, because there remains stage~\bbone\ work to do in its body.
So to proceed, the semantics must first search the body of the $\next$ for stage~\bbone\ subterms and reduce them in place.
This search process is called {\em speculation}.
In our example this means reducing all three projections to give,
\begin{lstlisting} 
prev{
  next{ 
    prev{next {2`f 20`}} 2`*` prev{next {2`f 20`}} 2`* 7` 
  }
}
\end{lstlisting}
Since all stage~\bbone\ subterms are reduced, the $\prev$s and $\next$s cancel to yield the final residual:
\begin{lstlisting} 
2`(f 20) * (f 20) * 7`
\end{lstlisting}
But now stage \bbtwo\ must compute the expensive function call \verb|f 20| twice! 
There are some systems, typically called {\em metaprogramming}, 
where this duplication would be considered intended behavior.
For our applications, we find the duplication undesirable and avoid it.

@@

More precisely, when evaluating the \verb|next|, we construct an explicit
substitution $\mathtt{[\hat y\mapsto f~20]}$ binding its old contents to the
fresh variable $\mathtt{\hat y}$. This substitution is placed at the top of the
containing $\prev$ block:
\begin{lstlisting} 
prev {
[2`yhat`|->2`f 20`]
  1`let x = (`next{2`yhat`}1`, 7) in`
  next{prev{1`#1 x`} 2`*` prev{1`#1 x`} 2`*`  hold{1`#2 x`}}
}
\end{lstlisting}
%As a convention, we render the new variable with a %stylish and fashionable
%hat.  
We proceed with stage~\bbone\ evaluation, which now duplicates the variable
$\mathtt{\hat y}$ rather than the expression \verb|f 20|.
\begin{lstlisting} 
prev {
[2`yhat`|->2`f 20`]
  next{
    prev{1`#1 (`next {2`yhat`}1`, 7)`} 2`*` 
    prev{1`#1 (`next {2`yhat`}1`, 7)`} 2`*`
    hold{1`#2 (`next {2`yhat`}1`, 7)`}
  }
}
\end{lstlisting}
We then speculate to reduce the projections and cancel the $\next$s and $\prev$s, giving:
\begin{lstlisting} 
prev {
[2`yhat`|->2`f 20`]
    next{ 2`yhat * yhat * 7` }
}
\end{lstlisting}
Once again, we lift the results of speculation into a substitution:
\begin{lstlisting} 
prev {
[2`yhat`|->2`f 20`]
[2`zhat`|->2`yhat * yhat * 7`]
    next{ 2`zhat` }
}
\end{lstlisting}
Finally, when canceling the last $\next$ and $\prev$, the semantics {\em reifies} the contained substitutions into let statements, yielding
\begin{lstlisting} 
2`let yhat = f 20 in
let zhat = yhat * yhat * 7 in zhat`
\end{lstlisting}

@@

The dynamics for \lang\ implements the stage \bbone\ evaluation algorithm
described above using three judgments: $\redonesym$ (evaluation), $\redtwosym$
(speculation), and the auxiliary $\reifysym$ (reification). 
Along with these are three judgments identifying forms of values:
$\pvalsym$ (partial values), $\ressym$ (residuals), 
and the auxiliary $\fcon$ (future context).
The interrelation of these judgments can be summarized as follows:
\begin{itemize}
\item Evaluation sends a stage~\bbone\ term to a partial value.
Because of the non-duplicating behavior, this sometimes produces explicit substitutions,
which are aggregated into a single {\em future context} that is also the output of evaluation.
\item Speculation sends a stage~\bbtwo\ term to a residual.
\item Reification sends a residual and a future context to another residual.  
\item Evaluation depends on speculation where stage~\bbone\ encapsulates stage~\bbtwo: $\next$. 
\item Speculation depends on evaluation where stage~\bbtwo\ encapsulates stage~\bbone: $\prev$ and $\pause$.
Reification is also used at these sites to handle the future context produced by evaluation.
\end{itemize}

Partial values are stage \bbone\ terms that have been fully
evaluated, but which may contain stage-\bbtwo\ variables wrapped in $\next$
blocks.  In the example above, 
\begin{lstlisting} 
1`(next {`2`yhat`1`},7)`
\end{lstlisting}
is a partial value.
Residuals ($\ressym$es) are \langTwo\ terms---stage \bbtwo\ terms whose
stage \bbone\ subexpressions have all been fully evaluated. In the example
above,
\begin{lstlisting} 
2`let zhat = yhat*yhat*7 in zhat`
\end{lstlisting}
is a residual,
These classifications are similar in that they both indicate that stage \bbone\ reduction is finished.
They are different in that ``partial value" is a classification of stage~\bbone\ terms,
while ``residual" is a classification of stage-\bbtwo\ terms.

@@

At $\next~e$, in contrast, $\redonesym$ must speculate into $e$ for stage~\bbone\ 
subexpressions to evaluate.  (See \ref{fig:diaSemanticsNP}.)
This is implemented by the $\redtwosym$ judgment, which takes a stage \bbtwo\ term to a residual. 
For the most part (\ref{fig:diaSemanticsSpec}), speculation simply recurs into
all subexpressions; at $\prev$, however, it resumes $\redonesym$ evaluation,
which produces a future context $\xi$ and (by canonical forms) an expression
$\next\{\hat y\}$. $\xi$ is then reified using $\reifysym$ into a series of let
bindings enclosing $\mathtt{\hat y}$.

@@

Within speculation lies a subtle---if perhaps unintuitive---feature.  
Observe that speculation will traverse into both branches of a stage-\bbtwo\ {\tt if} or {\tt case} 
statement in its search for stage~\bbone\ code. 
Thus the evaluation of that stage~\bbone\ code will occur {\em regardless of the eventual value of the predicate},
and so a term like 
\begin{lstlisting} 
next{
  2`if true 
  then `hold{1`3+4`} 
  2`else` prev{1`spin() (* loops forever *)`}
}
\end{lstlisting}
will fail to evaluate at stage \bbone.
This behavior is why the judgment is named ``speculation."
Although it may seem undesirable, this speculative behavior is a deliberate feature.
Indeed it is critical to the quickselect example, 
wherein all three branches of the $n$-$k$ comparison are evaluated at stage~\bbone.
As always, it is the programmer's responsibility to ensure that her staging annotations 
do not cause a program to loop forever.
